### 2020-04-14 thesis/content/chapter-ltd.tex
improved intro and conclusion
corrections

As per our definition of \ac{lta} from \cref{sec:ltag:definition}, the quality of an \ac{lta} formulation is determined by its chosen solution for the \ac{ltd} problem.
Since a comprehensive analysis of this problem would be beyond the scope of this thesis, we focus on the relation between \ac{ltd} and \acp{gcnn}.
The goal of this chapter is to answer the following two questions:
\begin{enumerate}
	\item \textit{Can a stack of graph convolution layers learn a decomposition function dynamically instead of using a static subtree decomposition?}
	\item \textit{What could be the foundation for such an ``\acs{ltd}-convolution'' layer?}
\end{enumerate}

@@

As we saw in \cref{sec:ltag:formulation:gcnn:conv}, the constituents of neighborhood convolutions are spanned by \ac{bfs} trees of depth $T$, where $T$ is the number of convolutional layers.
From this perspective the problem of learning constituents corresponds to learning a pruning operator on the branches of \ac{bfs} trees.
Such a pruning operator filters the edges that are traversed in each \ac{bfs} step.
There are two general edge filtering strategies:
\begin{enumerate}[label={\textbf{\arabic*.}}]
	\item \textbf{Edge prefiltering:}
		Here the edge filtering and convolution operations are performed in independent steps:
		First the edges of a given graph are filtered, then the convolution layers are applied to the filtered graph.
		This so-called \textit{edge prefiltering} strategy is illustrated in \cref{fig:ltd:subtree-filtering}.
		The main advantage of prefiltering is that it allows arbitrary combinations of edge filtering and convolution approaches.
		The main disadvantage is however that the same edges are removed in all \ac{bfs} subtrees.
		This restricts the expressive power of the learned decomposition function as shown in \cref{fig:ltd:dynamic-subtree-filtering}.
	\item \textbf{Dynamic edge filtering:}
		By filtering edges as part of the convolution operation itself, more flexible decompositions can be obtained.
		In the \textit{dynamic edge filtering} strategy the edge filter is part of the convolution operation and decides which neighbors of a given root node should be aggregated.
		Using this strategy a decomposition such as that shown in \cref{fig:ltd:dynamic-subtree-filtering} is learnable.
\end{enumerate}
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.8\linewidth]{gfx/ltd/subtree-filtering.pdf}
	\caption{Illustration of a pruned \ac{bfs} subtree when two edges are removed via prefiltering.}\label{fig:ltd:subtree-filtering}
\end{figure}
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.75\linewidth]{gfx/ltd/dynamic-subtree-filtering.pdf}
	\caption[A graph decomposition obtained via dynamic edge filtering which cannot be modeled by edge prefiltering.]{
		A graph decomposition obtained via dynamic edge filtering which cannot be modeled by edge prefiltering.
		Depending on the \ac{bfs} root node $v_{\mathrm{root}}$ different edges are removed.
		For $v_{\mathrm{root}} \in \textcolor{t_blue}{\{ v_3, v_4 \}}$ the edges $\{ (v_3, v_1), (v_3, v_2), (v_4, v_5), (v_4, v_6) \}$ are filtered out while for $v_{\mathrm{root}} \in \textcolor{t_red}{\{ v_1, v_2, v_5, v_6 \}}$ the edge $(v_3, v_4)$ is removed. % chktex 25
	}\label{fig:ltd:dynamic-subtree-filtering}
\end{figure}
Irrespective of the chosen edge filtering strategy, the decision whether to remove a given edge $e_{ij} = (v_i, v_j)$ or not has to be made based on relevant information about $e_{ij}$ and its surroundings.
One way to encode the information about $e_{ij}$ is via an edge feature vector $z_{ij} \in \mathbb{R}^d$ that is fed into the edge filter.

@@

The $k$-\acs{gnn} is a \ac{gcnn} inspired by the $k$-\ac{wl} algorithm, it convolves feature vectors of vertex $k$-multisets.
In this section we will compare its $k = 2$ variant with the 2-\ac{wl} algorithm.
The main difference between the two boils down to their notion of ``neighborhood''.
As already briefly mentioned in \cref{sec:related:gcr:nn}, 2-\acsp{gnn} define the neighbors of an edge $\textcolor{t_blue}{e_{ij}} = (v_i, v_j)$ to be the edges that are incident to either $v_i$ or $v_j$. % chktex 25
In 2-\acs{wl} on the other hand, the neighbors of $\textcolor{t_blue}{e_{ij}}$ are the edge pairs ${\left\{(\textcolor{t_red}{e_{il}}, \textcolor{t_darkgreen}{e_{lj}})\right\}}_{v_l \in \mathcal{V}_G}$, i.e.\ all possible paths of length two that start at $v_i$ and end at $v_j$. % chktex 21 chktex 25
This difference becomes clear when comparing the definition of convolution in 2-\acsp{gnn} with that of color refinement in 2-\acs{wl} (see \cfullref[ on ]{eq:related:kgnn-layer} and \cfullref[ on ]{defn:related:wlk-refine}):
\begin{alignat*}{4}
	\text{2-\acs{gnn}\footnotemark: } && Z_G^{(t)}[e_{ij}] &= \sigma\Biggl( \textcolor{t_blue}{Z_G^{(t-1)}[e_{ij}]} W^{(t)} + &&\left( \smashoperator[r]{\sum_{v_l \in \Gamma_G(v_j)}} \textcolor{t_red}{Z_G^{(t-1)}[e_{il}]} + \smashoperator[lr]{\sum_{v_l \in \Gamma_G(v_i)}} \textcolor{t_darkgreen}{Z_G^{(t-1)}[e_{lj}]} \right) W_{\Gamma}^{(t)} \Biggr) \\ % chktex 25
	\text{2-\acs{wl}: } && \chi_{G,2}^{(t)}(e_{ij}) &= h\Bigl(\textcolor{t_blue}{\chi_{G,2}^{(t-1)}(e_{ij})}, &&\ldblbrace (\textcolor{t_red}{\chi_{G,2}^{(t-1)}(e_{il})}, \textcolor{t_darkgreen}{\chi_{G,2}^{(t-1)}(e_{lj})})\, |\, v_l \in \mathcal{V}_G \rdblbrace\Bigr) % chktex 25
\end{alignat*}\footnotetext{
	To highlight the relation between 2-\acsp{gnn} and 2-\acs{wl}, a 2-\acs{gnn} definition that is not generally correct is shown here;
	for self-loops with $i = j$ it incorrectly sums the feature vectors of neighboring edges twice.
	The correct general formula uses a single sum over $v_l \in {\Gamma_G(v_j) \cup \Gamma_G(v_i)}$.
}\begin{figure}[ht]
	\centering
	\includegraphics[width=\linewidth]{gfx/ltd/2gnn-2wl-diff.pdf}
	\caption[Edge colorings on which 2-\acsp{gnn} and 2-\acs{wl} behave differently.]{
		Two edge colorings on which 2-\acsp{gnn} and 2-\acs{wl} behave differently.
		A 2-\acs{gnn} will refine the ``color vector'' of $e_{ij}$ to \colorlabel{t_blue}{D} for both initial colorings.
		2-\acs{wl} on the other hand differentiates both colorings by preserving the color tuple information.
	}\label{fig:ltd:2gnn-2wl-diff}
\end{figure}

We will now analyze what those different notions of neighborhood imply for the discriminative and computational power of 2-\acsp{gnn} in comparison to 2-\acs{wl}.
In the first step we show that the discriminative power of 2-\acsp{gnn} on all graphs $G \in \mathcal{G}$ is upper bounded by that of 1-\acs{wl} on the so-called \textit{edge neighborhood graphs} $G^{\mathcal{E}} \in \mathcal{G}^{\mathcal{E}}$.
\begin{defn}\label{defn:ltd:edge-graph}
	The \textit{edge neighborhood graph} of a given graph $G = (\mathcal{V}_G, \mathcal{E}_G)$ is defined as $G^{\mathcal{E}} \coloneqq (\mathcal{V}_{G^{\mathcal{E}}}, \mathcal{E}_{G^{\mathcal{E}}})$ with the vertices $\mathcal{V}_{G^{\mathcal{E}}} \coloneqq \{ \ldblbrace v, u \rdblbrace\, |\, (v, u) \in \mathcal{E}_G \lor v = u \}$ and the edges $\mathcal{E}_{G^{\mathcal{E}}} \coloneqq \left\{ (e, e') \in \mathcal{V}_{\mathcal{G}^{\mathcal{E}}}^2 |\, \left|e \cap e'\right| = 1 \right\}$. % chktex 21
\end{defn}
\begin{prop}\label{prop:ltd:2gnn-wl1-limit}
	The discriminative power of all 2-\acsp{gnn} $h_2$ is bounded by that of 1-\ac{wl} on edge neighborhood graphs, i.e.\ $\forall\, G, H \in \mathcal{G}: G^{\mathcal{E}} \mathrel{\simeq_1} H^{\mathcal{E}} \rightarrow h_2(G) = h_2(H)$. % chktex 21
\end{prop}
\begin{proof}
	By \cref{defn:ltd:edge-graph} it holds that $\forall e_{ij} \in \mathcal{E}_G: {\Gamma_G(v_i) \cup \Gamma_G(v_j)} = \Gamma_{G^{\mathcal{E}}}(e_{ij})$.
	Therefore the 2-\acs{gnn} convolution operation defined in \cfullref[ on ]{eq:related:kgnn-layer} can be rewritten as a vertex neighborhood convolution operator
	\begin{align*}
		Z_G^{(t)}[e] = \sigma\left( Z_G^{(t-1)}[e] W^{(t)} + \smashoperator[lr]{\sum_{e' \in \Gamma_{G^{\mathcal{E}}}(e)}} Z_G^{(t-1)}[e'] W_{\Gamma}^{(t)} \right)
		\text{.}
	\end{align*}
	\Cref{prop:ltd:2gnn-wl1-limit} then follows from \cref{prop:related:gcnn-wl1-limit}~(\cpageref{prop:related:gcnn-wl1-limit}).
\end{proof}
\begin{lem}\label{lem:ltd:wl1-regular-edge-neighbor-limit}
	1-\ac{wl} cannot distinguish the edge neighborhood graphs $G^{\mathcal{E}}$ and $H^{\mathcal{E}}$ of any pair of $d$-regular graphs $G$ and $H$ with $n$ vertices.
\end{lem}
\begin{proof}
	Let $G$ and $H$ be two $d$-regular graphs of size $n$.
	Their corresponding edge neighborhood graphs $G^{\mathcal{E}}$ and $H^{\mathcal{E}}$ both have $n^{\mathcal{E}} = n + \frac{nd}{2}$ vertices.
	$n$ of those edge neighborhood vertices correspond to the vertices of $G$ and $H$ respectively, we will refer to them as \textit{loop vertices} $L_G$/$L_H$.
	The remaining $\frac{nd}{2}$ edge neighborhood vertices correspond to the edges of $G$ and $H$, we will refer to them as \textit{edge vertices} $E_G$/$E_H$.

@@

As described in \cref{sec:related:character:wl} the discriminative power of a model is, by itself, not necessarily relevant for real-world graph datasets.
In \cfullref[ on ]{prop:related:wl2-cycle-count} we saw however that 2-\acs{wl} not only has a higher discriminative power than 1-\acs{wl} but also that it is able to count the number of $m$-cycles in a given graph for all $m \leq 7$.
This is relevant because cycle counts are a commonly used metric in real-world domains such as social network and molecular structure analysis~\cite{Milo2002}\cite{Newman2003}\cite{Welser2007}\cite{Adamson1973}\cite{Kekule1866}.
To conclude this section we now show that 2-\acsp{gnn} not only have a lower discriminative power than 2-\acs{wl} but are also unable to detect cycles.
\begin{prop}
	2-\acsp{gnn} cannot detect $m$-cycles for all $m \geq 3$.
\end{prop}
\begin{proof}
	Let $n$ be the \ac{lcm} of $3$ and some $m > 3$.
	We define $c_3 \coloneqq \frac{n}{3}$ and $c_m \coloneqq \frac{n}{m}$.
	Based on that we define the following two graphs:
	Let $G_3$ be a graph consisting of $c_3$ disconnected cycles of length $3$, analogously let $G_m$ be a graph consisting of $c_m$ disconnected cycles of length $m$.
	Since both $G_3$ and $G_m$ are 2-regular and have the size $n$, any 2-\ac{gnn} $h_2: \mathcal{G} \to \mathcal{Y}$ must map both of them to the same $y \in \mathcal{Y}$ by \cref{prop:ltd:2gnn-regular-limit}.

	Let us assume that $h_2$ is able to detect cycles of length $3$, i.e.\ triangles.
	By \cfullref[ on ]{defn:related:wl-count-detect} this would imply that there is a function $g: \mathcal{Y} \to \{ 0, 1\}$ s.t.\ $g(h_2(G_3)) = g(y) = 1$ and $g(h_2(G_m)) = g(y) = 0\ \lightning$.
	Conversely, assuming that $h_2$ is able to detect cycles of length $m > 3$, we obtain the contradiction $g(h_2(G_3)) = g(y) = 0$  and $g(h_2(G_m)) = g(y) = 1\ \lightning$.
	This concludes the proof.
\end{proof}

In this section we compared 2-\acsp{gnn} with the 2-\acs{wl} algorithm and found that they have a significantly lower discriminative and computational power than 2-\acs{wl}.
Motivated by those limitations we describe a novel graph convolution operator which is closer to 2-\acs{wl} in the following section.

@@

We have seen that the main difference between 2-\acsp{gnn} and 2-\acs{wl} is their notion of ``neighborhood''.
In this section we describe a novel convolution operator which uses edge tuple neighbors just like 2-\acs{wl} to overcome the limitations of 2-\acsp{gnn}.
This will be done in three steps:
\begin{enumerate*}
	\item We begin by formally defining the 2-\acs{wl} convolution operator.
	\item Then the discriminative and computational power of this operator are analyzed.
	\item Lastly we describe how 2-\acs{wl} convolutions can be implemented efficiently on a modern \ac{gpgpu}.
\end{enumerate*}

@@

Similar to the 2-LWL kernel described in \cref{sec:related:gcr:kernel}, our 2-\acs{wl} convolution operator reduces the computational cost of 2-\acs{wl} via two simplifications:
\begin{enumerate}[label=\textbf{\arabic*.}]
	\item \textbf{2-multisets:}
		Since we assume that graphs are undirected, $e_{ij}$ and $e_{ji}$ have identical feature vectors $x[e_{ij}] = x[e_{ji}] \in \mathcal{X}_{\mathcal{E}}$ and the same 2-\acs{wl} neighborhood.
		Instead of refining/convolving the feature vectors of 2-tuples $(v_i, v_j)$ we can therefore refine/convolve the feature vectors of 2-multisets $\ldblbrace v_i, v_j \rdblbrace$.
		This halves the number of feature vectors without affecting the discriminative or computational power of the 2-\acs{wl} convolution operator.
		To simplify the notation we assume that $e_{ij} = e_{ji} = \ldblbrace v_i, v_j \rdblbrace$ in the rest of this section.
	\item \textbf{Neighborhood localization:}
		Using the 2-multiset simplification, the original 2-\acs{wl} algorithm refines the color of all multisets $e_{ij} \in \mathcal{V}_G^2$ by hashing its current color and the colors of all neighbors ${\{ \ldblbrace e_{il}, e_{lj} \rdblbrace \}}_{v_l \in \mathcal{V}_G}$.
		This means that the time complexity of a single refinement step is $\mathcal{O}(n^3)$ for $n \coloneqq \left| \mathcal{V}_G \right|$ which quickly becomes infeasible for large graphs.

		To address this issue we reduce both the number of colored edges as well as the number of neighbors of each edge.
		This is achieved by only considering the edges that are part of the so-called \textit{$r$-th power of a given graph $G$} where $r \in \mathbb{N}$ is the freely choosable \textit{neighborhood radius}.
		\begin{defn}
			The \textit{$r$-th power of a graph $G$} is defined as
			\begin{align*}
				G^r \coloneqq \left(\mathcal{V}_G, \left\{ e_{ij} \in \mathcal{V}_G^2\, |\, d_{\mathrm{SP}, G}(v_i, v_j) \leq r \right\}\right) % chktex 21
			\end{align*}
			where $d_{\mathrm{SP}, G}(v_i, v_j)$ is the length of the shortest path between $v_i$ and $v_j$ in $G$.
			The distance of a vertex $v_i \in \mathcal{V}_G$ to itself is defined as $d_{\mathrm{SP}, G}(v_i, v_i) \coloneqq 0$.
			Note that $G^1$ does not generally equal $G$ because $G^1$ has self-loop edges $e_{ii} \in \mathcal{E}_{G^1}$ at all vertices.
		\end{defn}
		For the neighborhood radius $r = 1$ only the self-loop edges ${\{ e_{ii} \}}_{v_i \in \mathcal{V}_G}$ and the edges $\mathcal{E}_{G}$ are considered;
		for $r > 1$ edges between indirectly connected vertices are considered as well.
		\Cref{fig:ltd:neighborhood-radius} illustrates this for varying neighborhood radii.
		Through the reduction of the considered edges, the neighbors of each $e_{ij} \in \mathcal{E}_{G^r}$ are in turn reduced to the common $r$-neighbors of $v_i$ and $v_j$, i.e.\ $\{ \ldblbrace e_{il}, e_{lj} \rdblbrace\, |\, v_l \in {\Gamma_{G^r}(v_i) \cap \Gamma_{G^r}(v_j)} \}$. % chktex 21
		\begin{figure}[t]
			\centering
			\includegraphics[width=0.9\linewidth]{gfx/ltd/neighborhood-radius.pdf}
			\caption[Illustration of the powers of the six-cycle graph.]{
				Illustration of the powers of the six-cycle graph for varying $r$.
				The self-loop edge at each of the vertices is not explicitly shown.
				For $r = 3$ all possible edges between the six vertices will be considered just as in the original 2-\acs{wl} algorithm.
			}\label{fig:ltd:neighborhood-radius}
		\end{figure}

@@

Based on the 2-multiset and the neighborhood localization simplifications we can now define the 2-\ac{wl} convolution operator and the corresponding so-called \textit{2-\acs{wl}-\acs{gnn}}.

>>>>

As per our definition of \ac{lta} from \cref{sec:ltag:definition}, the quality of an \ac{lta} formulation is determined by its chosen solution to the \ac{ltd} problem.
Since a comprehensive analysis of this problem would be beyond the scope of this thesis, we focus on the relation between \ac{ltd} and \acp{gcnn}.
The goal of this chapter is to answer the following two questions:
\begin{enumerate}
	\item \textit{Can a stack of graph convolution layers learn a decomposition function dynamically instead of using a static subtree decomposition?}
	\item \textit{What could be the foundation for such an ``\acs{ltd}-convolution'' layer?}
\end{enumerate}

@@

As we saw in \cref{sec:ltag:formulation:gcnn:conv}, the constituents of neighborhood convolutions are spanned by \ac{bfs} trees of depth $T$, where $T$ is the number of convolutional layers.
From this perspective the problem of learning constituents corresponds to learning a pruning operator on the branches of \ac{bfs} trees.
Such a pruning operator filters the edges that are traversed in each \ac{bfs} step.
There are two general edge filtering strategies:
\begin{enumerate}[label={\textbf{\arabic*.}}]
	\item \textbf{Edge prefiltering:}
		Here the edge filtering and convolution operations are performed in independent steps:
		First the edges of a given graph are filtered, then the convolution layers are applied to the filtered graph.
		This so-called \textit{edge prefiltering} strategy is illustrated in \cref{fig:ltd:subtree-filtering}.
		The main advantage of prefiltering is that it allows arbitrary combinations of edge filtering and convolution approaches.
		The main disadvantage is however that the same edges are removed in all \ac{bfs} subtrees.
		This restricts the expressive power of the learned decomposition function as shown in \cref{fig:ltd:dynamic-subtree-filtering}.
	\item \textbf{Dynamic edge filtering:}
		By filtering edges as part of the convolution operation itself, more flexible decompositions can be obtained.
		In the \textit{dynamic edge filtering} strategy the edge filter is part of the convolution operation and decides which neighbors of a given root node should be aggregated.
		Using this strategy a decomposition such as that shown in \cref{fig:ltd:dynamic-subtree-filtering} is learnable.
\end{enumerate}
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.8\linewidth]{gfx/ltd/subtree-filtering.pdf}
	\caption{Illustration of a pruned \ac{bfs} subtree when two edges are removed via prefiltering.}\label{fig:ltd:subtree-filtering}
\end{figure}
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.75\linewidth]{gfx/ltd/dynamic-subtree-filtering.pdf}
	\caption[A graph decomposition obtained via dynamic edge filtering which cannot be modeled by edge prefiltering.]{
		A graph decomposition obtained via dynamic edge filtering which cannot be modeled by edge prefiltering.
		Depending on the \ac{bfs} root node $v_{\mathrm{root}}$, different edges are removed.
		For $v_{\mathrm{root}} \in \textcolor{t_blue}{\{ v_3, v_4 \}}$ the edges $\{ (v_3, v_1), (v_3, v_2), (v_4, v_5), (v_4, v_6) \}$ are filtered out while for $v_{\mathrm{root}} \in \textcolor{t_red}{\{ v_1, v_2, v_5, v_6 \}}$ the edge $(v_3, v_4)$ is removed. % chktex 25
	}\label{fig:ltd:dynamic-subtree-filtering}
\end{figure}
Irrespective of the chosen edge filtering strategy, the decision whether to remove a given edge $e_{ij} = (v_i, v_j)$ or not has to be made based on relevant information about $e_{ij}$ and its surroundings.
One way to encode the information about $e_{ij}$ is via an edge feature vector $z_{ij} \in \mathbb{R}^d$ that is fed into the edge filter.

@@

The $k$-\acs{gnn} is a \ac{gcnn} inspired by the $k$-\ac{wl} algorithm, it convolves feature vectors of vertex $k$-multisets.
In this section we will compare its $k = 2$ variant with the 2-\ac{wl} algorithm.
The main difference between the two boils down to their notion of ``neighborhood''.
As already briefly mentioned in \cref{sec:related:gcr:nn}, 2-\acsp{gnn} define the neighbors of an edge $\textcolor{t_blue}{e_{ij}} = (v_i, v_j)$ to be the edges that are incident to either $v_i$ or $v_j$. % chktex 25
In 2-\acs{wl} on the other hand, the neighbors of $\textcolor{t_blue}{e_{ij}}$ are the edge pairs ${\left\{(\textcolor{t_red}{e_{il}}, \textcolor{t_darkgreen}{e_{lj}})\right\}}_{v_l \in \mathcal{V}_G}$, i.e.\ all possible walks of length two that start at $v_i$ and end at $v_j$. % chktex 21 chktex 25
This difference becomes clear when comparing the definition of convolution in 2-\acsp{gnn} with that of color refinement in 2-\acs{wl} (see \cfullref[ on ]{eq:related:kgnn-layer} and \cfullref[ on ]{defn:related:wlk-refine}):
\begin{alignat*}{4}
	\text{2-\acs{gnn}\footnotemark: } && Z_G^{(t)}[e_{ij}] &= \sigma\Biggl( \textcolor{t_blue}{Z_G^{(t-1)}[e_{ij}]} W^{(t)} + &&\left( \smashoperator[r]{\sum_{v_l \in \Gamma_G(v_j)}} \textcolor{t_red}{Z_G^{(t-1)}[e_{il}]} + \smashoperator[lr]{\sum_{v_l \in \Gamma_G(v_i)}} \textcolor{t_darkgreen}{Z_G^{(t-1)}[e_{lj}]} \right) W_{\Gamma}^{(t)} \Biggr) \\ % chktex 25
	\text{2-\acs{wl}: } && \chi_{G,2}^{(t)}(e_{ij}) &= h\Bigl(\textcolor{t_blue}{\chi_{G,2}^{(t-1)}(e_{ij})}, &&\ldblbrace (\textcolor{t_red}{\chi_{G,2}^{(t-1)}(e_{il})}, \textcolor{t_darkgreen}{\chi_{G,2}^{(t-1)}(e_{lj})})\, |\, v_l \in \mathcal{V}_G \rdblbrace\Bigr) % chktex 25
\end{alignat*}\footnotetext{
	To highlight the relation between 2-\acsp{gnn} and 2-\acs{wl}, a 2-\acs{gnn} definition that is not generally correct is shown here;
	for self-loops with $i = j$ it incorrectly sums the feature vectors of neighboring edges twice.
	The correct general formula uses a single sum over $v_l \in {\Gamma_G(v_j) \cup \Gamma_G(v_i)}$.
}\begin{figure}[ht]
	\centering
	\includegraphics[width=\linewidth]{gfx/ltd/2gnn-2wl-diff.pdf}
	\caption[Edge colorings on which 2-\acsp{gnn} and 2-\acs{wl} behave differently.]{
		Two edge colorings on which 2-\acsp{gnn} and 2-\acs{wl} behave differently.
		A 2-\acs{gnn} will refine the ``color vector'' of $e_{ij}$ to \colorlabel{t_blue}{D} for both initial colorings.
		2-\acs{wl} on the other hand differentiates both colorings by preserving the color tuple information.
	}\label{fig:ltd:2gnn-2wl-diff}
\end{figure}

We will now analyze what those different notions of neighborhood imply for the discriminative and computational power of 2-\acsp{gnn} in comparison to 2-\acs{wl}.
In the first step we show that the discriminative power of 2-\acsp{gnn} on all graphs $G \in \mathcal{G}$ is upper bounded by that of 1-\acs{wl} on the so-called \textit{edge neighborhood graphs} $G^{\mathcal{E}} \in \mathcal{G}^{\mathcal{E}}$.
\begin{defn}\label{defn:ltd:edge-graph}
	The \textit{edge neighborhood graph} of a given graph $G = (\mathcal{V}_G, \mathcal{E}_G)$ is defined as $G^{\mathcal{E}} \coloneqq (\mathcal{V}_{G^{\mathcal{E}}}, \mathcal{E}_{G^{\mathcal{E}}})$ with the vertices $\mathcal{V}_{G^{\mathcal{E}}} \coloneqq \{ \ldblbrace v, u \rdblbrace\, |\, (v, u) \in \mathcal{E}_G \lor v = u \}$ and the edges $\mathcal{E}_{G^{\mathcal{E}}} \coloneqq \left\{ (e, e') \in \mathcal{V}_{\mathcal{G}^{\mathcal{E}}}^2 |\, \left|e \cap e'\right| = 1 \right\}$. % chktex 21
\end{defn}
\begin{prop}\label{prop:ltd:2gnn-wl1-limit}
	The discriminative power of all 2-\acsp{gnn} $h_2$ is bounded by that of 1-\ac{wl} on edge neighborhood graphs, i.e.\ $\forall\, G, H \in \mathcal{G}: G^{\mathcal{E}} \mathrel{\simeq_1} H^{\mathcal{E}} \rightarrow h_2(G) = h_2(H)$. % chktex 21
\end{prop}
\begin{proof}
	By \cref{defn:ltd:edge-graph}, it holds that $\forall e_{ij} \in \mathcal{E}_G: {\Gamma_G(v_i) \cup \Gamma_G(v_j)} = \Gamma_{G^{\mathcal{E}}}(e_{ij})$.
	Therefore the 2-\acs{gnn} convolution operation defined in \cfullref[ on ]{eq:related:kgnn-layer} can be rewritten as a vertex neighborhood convolution operator
	\begin{align*}
		Z_G^{(t)}[e] = \sigma\left( Z_G^{(t-1)}[e] W^{(t)} + \smashoperator[lr]{\sum_{e' \in \Gamma_{G^{\mathcal{E}}}(e)}} Z_G^{(t-1)}[e'] W_{\Gamma}^{(t)} \right)
		\text{.}
	\end{align*}
	\Cref{prop:ltd:2gnn-wl1-limit} then follows from \cref{prop:related:gcnn-wl1-limit}~(\cpageref{prop:related:gcnn-wl1-limit}).
\end{proof}
\begin{lem}\label{lem:ltd:wl1-regular-edge-neighbor-limit}
	1-\ac{wl} cannot distinguish the edge neighborhood graphs $G^{\mathcal{E}}$ and $H^{\mathcal{E}}$ of any pair of $d$-regular graphs $G$ and $H$ with $n$ vertices.
\end{lem}
\begin{proof}
	Let $G$ and $H$ be two $d$-regular graphs of size $n$.
	Their corresponding edge neighborhood graphs $G^{\mathcal{E}}$ and $H^{\mathcal{E}}$ both have $n^{\mathcal{E}} = n + \frac{nd}{2}$ vertices.
	$n$ of those edge neighborhood vertices correspond to the vertices of $G$ and $H$ respectively, we will refer to them as \textit{loop vertices} $L_G$/$L_H$.
	The remaining $\frac{nd}{2}$ edge neighborhood vertices correspond to the edges of $G$ and $H$, we will refer to them as \textit{edge vertices} $E_G$/$E_H$.

@@

As described in \cref{sec:related:character:wl}, the discriminative power of a model is, by itself, not necessarily relevant for real-world \ac{gcr} problems.
In \cfullref[ on ]{prop:related:wl2-cycle-count} we saw however that 2-\acs{wl} not only has a higher discriminative power than 1-\acs{wl} but also that it is able to count the number of $m$-cycles in a given graph for all $m \leq 7$.
This is relevant because cycle counts are a commonly used metric in real-world domains such as social network and molecular structure analysis~\cite{Milo2002}\cite{Newman2003}\cite{Welser2007}\cite{Adamson1973}\cite{Kekule1866}.
To conclude this section, we now show that 2-\acsp{gnn} not only have a lower discriminative power than 2-\acs{wl} but are also unable to detect cycles.
\begin{prop}\label{prop:ltd:2gnn-cycle-limit}
	2-\acsp{gnn} cannot detect $m$-cycles for all $m \geq 3$.
\end{prop}
\begin{proof}
	Let $n$ be the \ac{lcm} of $3$ and some $m > 3$.
	We define $c_3 \coloneqq \frac{n}{3}$ and $c_m \coloneqq \frac{n}{m}$.
	Based on that, we define the following two graphs:
	Let $G_3$ be a graph consisting of $c_3$ disconnected cycles of length $3$, analogously let $G_m$ be a graph consisting of $c_m$ disconnected cycles of length $m$.
	Since both $G_3$ and $G_m$ are 2-regular and have the size $n$, any 2-\ac{gnn} $h_2: \mathcal{G} \to \mathcal{Y}$ must map both of them to the same $y \in \mathcal{Y}$ by \cref{prop:ltd:2gnn-regular-limit}.

	Let us assume that $h_2$ is able to detect cycles of length $3$, i.e.\ triangles.
	By \cfullref[ on ]{defn:related:wl-count-detect}, this would imply that there is a function $g: \mathcal{Y} \to \{ 0, 1\}$ s.t.\ $g(h_2(G_3)) = g(y) = 1$ and $g(h_2(G_m)) = g(y) = 0\ \lightning$.
	Conversely, assuming that $h_2$ is able to detect cycles of length $m > 3$, we obtain the contradiction $g(h_2(G_3)) = g(y) = 0$  and $g(h_2(G_m)) = g(y) = 1\ \lightning$.
	This concludes the proof.
\end{proof}

In this section we compared 2-\acsp{gnn} with the 2-\acs{wl} algorithm and found that they have a significantly lower discriminative and computational power than 2-\acs{wl}.
Motivated by those limitations we describe a novel graph convolution operator, which is closer to 2-\acs{wl}, in the following section.

@@

We have seen that the main difference between 2-\acsp{gnn} and 2-\acs{wl} is their notion of ``neighborhood''.
In this section we describe a novel convolution operator which uses edge tuple neighbors just like 2-\acs{wl} to overcome the limitations of 2-\acsp{gnn}.
This will be done in three steps:
\begin{enumerate*}[label={\circled{\small\arabic*}}]
	\item We begin by formally defining the 2-\acs{wl} convolution operator.
	\item Then the discriminative and computational power of this operator is analyzed.
	\item Lastly, we describe how 2-\acs{wl} convolutions can be implemented efficiently on a modern \ac{gpgpu}.
\end{enumerate*}

@@

Similar to the 2-LWL kernel described in \cref{sec:related:gcr:kernel}, our 2-\acs{wl} convolution operator reduces the computational cost of 2-\acs{wl} via two simplifications:
\begin{enumerate}[label=\textbf{\arabic*.}]
	\item \textbf{2-multisets:}
		Since we assume that graphs are undirected, $e_{ij}$ and $e_{ji}$ have identical feature vectors $x[e_{ij}] = x[e_{ji}] \in \mathcal{X}_{\mathcal{E}}$ and the same 2-\acs{wl} neighborhood.
		Instead of refining/convolving the feature vectors of 2-tuples $(v_i, v_j)$, we can therefore refine/convolve the feature vectors of 2-multisets $\ldblbrace v_i, v_j \rdblbrace$.
		This halves the number of feature vectors without affecting the discriminative or computational power of the 2-\acs{wl} convolution operator.
		To simplify the notation we assume that $e_{ij} = e_{ji} = \ldblbrace v_i, v_j \rdblbrace$ in the rest of this section.
	\item \textbf{Neighborhood localization:}
		Using the 2-multiset simplification, the original 2-\acs{wl} algorithm refines the color of all multisets $e_{ij} \in \mathcal{V}_G^2$ by hashing its current color and the colors of all neighbors ${\{ \ldblbrace e_{il}, e_{lj} \rdblbrace \}}_{v_l \in \mathcal{V}_G}$.
		This means that the time complexity of a single refinement step is $\mathcal{O}(n^3)$ for $n \coloneqq \left| \mathcal{V}_G \right|$, which quickly becomes infeasible for large graphs.

		To address this issue we reduce both the number of colored edges as well as the number of neighbors of each edge.
		This is achieved by only considering the edges that are part of the so-called \textit{$r$-th power of a given graph $G$} where $r \in \mathbb{N}$ is the freely choosable \textit{neighborhood radius}.
		\begin{defn}
			The \textit{$r$-th power of a graph $G$} is defined as
			\begin{align*}
				G^r \coloneqq \left(\mathcal{V}_G, \left\{ e_{ij} \in \mathcal{V}_G^2\, |\, d_{\mathrm{SP}, G}(v_i, v_j) \leq r \right\}\right) % chktex 21
			\end{align*}
			where $d_{\mathrm{SP}, G}(v_i, v_j)$ is the length of the shortest path between $v_i$ and $v_j$ in $G$.
			The distance of a vertex $v_i \in \mathcal{V}_G$ to itself is defined as $d_{\mathrm{SP}, G}(v_i, v_i) \coloneqq 0$.
			Note that $G^1$ does not generally equal $G$ because $G^1$ has self-loop edges $e_{ii} \in \mathcal{E}_{G^1}$ at all vertices.
		\end{defn}
		For the neighborhood radius $r = 1$, only the self-loop edges ${\{ e_{ii} \}}_{v_i \in \mathcal{V}_G}$ and the edges $\mathcal{E}_{G}$ are considered;
		for $r > 1$, edges between indirectly connected vertices are considered as well.
		\Cref{fig:ltd:neighborhood-radius} illustrates this for varying neighborhood radii.
		Through the reduction of the considered edges, the neighbors of each $e_{ij} \in \mathcal{E}_{G^r}$ are in turn reduced to the common $r$-neighbors of $v_i$ and $v_j$, i.e.\ $\{ \ldblbrace e_{il}, e_{lj} \rdblbrace\, |\, v_l \in {\Gamma_{G^r}(v_i) \cap \Gamma_{G^r}(v_j)} \}$. % chktex 21
		\begin{figure}[t]
			\centering
			\includegraphics[width=0.9\linewidth]{gfx/ltd/neighborhood-radius.pdf}
			\caption[Illustration of the powers of the six-cycle graph.]{
				Illustration of the powers of the six-cycle graph for varying $r$.
				The self-loop edge at each of the vertices is not explicitly shown.
				For $r = 3$ all possible edges between the six vertices will be considered, just as in the original 2-\acs{wl} algorithm.
			}\label{fig:ltd:neighborhood-radius}
		\end{figure}

@@

Based on the 2-multiset and the neighborhood localization simplifications, we now define the 2-\ac{wl} convolution operator and the corresponding so-called \textit{2-\acs{wl}-\acs{gnn}}.
